Monte Carlo estimation of the probability of causal contacts between communicating civilisations

### 2020-07-17 submit_v05/manuscript.tex
final version

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{graphics}
\usepackage{algorithmic}
\usepackage{subfig}
\usepackage{url}

\usepackage{boites}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{condition}{Condition}

@@

\newcommand{\awakening}{A event }
\newcommand{\contact}{C event }
\newcommand{\blackout}{B event }
\newcommand{\doomsday}{D event }

\newcommand{\aawakening}{Awakening event }
\newcommand{\ccontact}{Contact event }
\newcommand{\bblackout}{Blackout event }
\newcommand{\ddoomsday}{Doomsday event }

\newcommand{\awakenings}{A events }
\newcommand{\contacts}{C events }
\newcommand{\blackouts}{B events }
\newcommand{\doomsdays}{D events }

\newcommand{\ff}{{\color{red}\rule{10pt}{10pt}}}

%\newcommand{\ffn}[1]{ \colorbox{red}{FIG {#1}} }    
\newcommand{\ffn}[1]{}

%\newcommand{\ttn}[1]{ \colorbox{green}{TABLE {#1}} }    
\newcommand{\ttn}[1]{}

@@

%\JELclassification{XX; XX}
%\MSCcodes{XX; XX}

@@

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilisations in the Milky Way. 
%
The equation quantifies the number of civilisations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimated
values are revised continually. 
%
A comprehensive review and an analysis of each term of the equation are
presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,
barlow_galactic_2012, forgan_galactic_2016, anchordoqui_is_2017,
Sotos_biotechnology_2019, carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by
astrophysical phenomena that makes life difficult to develop
\citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life,
alternative scenarios have also been discussed
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilisations may play an important role in determining how difficult it
would be to obtain evidence for other inhabited worlds.
%
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citet{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the unknown abundance of
civilisations in the Galaxy has been organised around the factors in
the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it less suitable to a formal
study with the purpose of defining searching strategies or computing
the estimated number of extraterrestrial intelligences.
%
A number of studies propose alternative formalisms for the estimation
of the likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework
for a joint analysis of the Drake equation and the Fermi paradox,
concluding that for sufficiently long-lived civilisations,
colonisation is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.

@@

On this topic a number of works consider numerical simulations
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017,
ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations (like the fraction of stars with planets), it
depends on the definition of unknown or uncertain parameters required
to carry out the simulations.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a two--dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6~kpc and a peak radius of 10~kpc.
%
These simulations represent the spread of intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI searches
from a novel point of view.

% INTRODUCTION OF *THIS* WORK

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of
life, the development of intelligence or the formation of habitable
planets according to stellar type.
%
Instead, we assume monoparametric function families to model the
appearance of points in the disk of the Galaxy and over the time,
which we call \textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that
are able to receive and emit signals with perfect efficiency in all
directions.
%
Then we analyse the network of causally connected nodes limited to
have a maximum separation representing the maximum distance a signal
can travel with an intensity above a fixed threshold.
%
A node is causally connected to other node if it is within their light
cone.
%
We adopt as a definition that a light cone is the region in spacetime
within the surface generated by the light emanating from a given point
in space in a given period of time.
%
For example, two nodes separated by 100 light years which appear with
a difference in time of 80 years and last 50 years each, will be
causally connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the
first node will never acknowledge the existence of the second.
%
An scheme of this example is shown if Fig.~\ref{F_example}.
%
This simple tought experiment exemplify the importance of the time
variable to analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean
separation between the appearance of nodes, the mean lifetime of the
nodes, and the maximum separation to allow an effective causal
contact.
%
The model presented here comprises the stochastic networks of
constrained causally connected nodes, together with the three
parameters and additional hypotheses.
%
Hereafter, we refer to this model as the SC3Net model, standing for
''Stochastic Constrained Causally Connected Network``.

@@

\ffn{1} Space--time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen to signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen to signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

As part of this benchmark, we assume that the capacity to emit and receive
signals occur at the same time.
%
Although there are several reasons to think that this could not be the case, at
large time scales it can be considered that both abilities occur roughly at
coincident epochs.
%
In the ideal setup, this would be equivalent to nodes that send messages
isotropically and scan the local skies on all directions with a perfect
eficiency.
%
Another essential assumption is that all \cetis{} use the same signal power, so
that there is a maximum distance out to which it can be detected.
%
In such a system we compute probabilities of a random node making contact with
another node, i.e., they are not specific for the case of contacts with the
Earth.
%
Regarding the extent of the signals, we know that the distance from which a
signal from Earth could be detected using the current technology is about few
parsecs, given that the signal was sent to a specific direction.
%
It is straightforward to propose and implement a distribution of maximum
distances, although this would increase the model complexity at the cost of a
larger uncertainty.

@@

\hline
\end{tabular}
\caption{Definition of independent variables and adopted values for 
   fixed parameters 
   that are part of the simulation.  Variable parameters define the
   spatial and temporal structure
   of the process and the maximum reach of the messages.}
\label{T_simu_hypotheses}
\end{table*}

@@

% hearsay 
%
In order to make a reproducible project, we developed the tools that
allow to run the simulations and obtain the results shown in this
work.
%
The simulations were implemented on a Python-3 code, dubbed \hs{}
(Lares et al., in preparation), which is publicly accessible through
the GitHub platform\footnote{https://github.com/mlares/hearsay} under
the MIT-license.
%
The project is in the process of registration with the ‘’Astrophysics
Source Code Library’’ \citep[ASCL, ][]{2015JORS....3E..15A,
2020ASPC..522..731A}
%
From the viewpoint of a user, \hs{} is an object-oriented package that
exposes the main functionalities as classes and methods.
%
The code fulfills standar quality assurance metrics, that account for
testing, style, documentation and coverage.
%
In the configuration step, the user prepares the set of simulation
parameters through an initialisation file.
%
Since configuration files and simulation results are persistent, it is
straightforward to keep track of different experiments and the
experiments can be revisited easily.
%
An in-depth description of the methods can be found in the
documentation, which is automatically generated from \hs{} docstrings
and made public in the read-the-docs service
\footnote{https://hearsay.readthedocs.io/en/latest/}.
%
Since the simulation setup is configurable, the time required for a
simulation to complete depends on the simulation parameters.
%
It also depends on the hardware that is used to make the run, and on
whether the parallel option is set.
%
However, it is simple to make a local run of limited versions of the
experiments to have a sense of the time required by the code to
complete the simulations.
%
More information on this can be found on the documentation of the
software.

@@

We implemented the simulation of a regular grids of models varying over
the parameter space, which covers 5204 models.
%
For each model, we simulated several realizations with different random
seeds, adding up a total of 
158546 simulation runs.
%
The number of random realzations varies from one (for the densest populated models)
up to 50 (for the sparsely populated models).
%
The parameters for the temporal aspects of the simulation (the mean
waiting time for the next awakening, $\tau_{\rm a}$, and the
mean lifetime, $\tau_{\rm s}$) cover the ranges
10$^2$-10$^6$~yr, with 
two linear partitions of 21 values (10$^2$-40500~yr) and 51 values
($5\cdot10^5$-10$^6$~yr)
for each parameter.
%
This partition was chosen after the requirement of the software to take linear bins, 
aiming at a better sampling of the low $\tau_{\rm a}$ and low
$\tau_{\rm s}$ region of the parameter space.
%
For the D$_{\rm max}$ parameter, we take the 
values 100~pc, 500~pc, 1~kpc, 5~kpc and 10~kpc.
%
\ttn{1}
%
In the Table \ref{T_simu_hypotheses} we show the three variable
parameters, the ranges of their values and the number of bins that
have been explored in the numerical experiments.
%
We also show the set of fixed parameters that take part in the simulation,
their values and the hypotheses that they represent.

As a product of the simulations, several quantities can be obtained.
%
Some of them are directly derived from the discrete events, namely,
the ID of emitting and receiving \cetis{} and the position in the
galaxy.
%
We also save the times of each of the events that are relevant to keep
track of the number of \cetis{} for each simulation, i.e., the times
of the four types of events.
%
The times of C-type and B-type events are used to derive the number of
contacts.
%
We also obtain quantities that represent the properties of the
\cetis{}, for example the total time elapsed between the A-type and
the D-type of each node, which represent their corresponding
lifetimes.
%
The time span of a \ceti{} listening another or being listened by
another node can also be derived by keepig track of the times of the
events in a simulation.
%
This way we can also compute the distribution in the simulated galaxy
of \cetis{} that reach contact, the distributions of the waiting times
until the first contact or the distributions of waiting times until
the next contact.
%
The following properties of the population of \cetis{} can also be
derived:
%
the fraction of the lifetime a node is listening to at least another
node (i.e., within their light cone),
%
the age of contacted \cetis{} at first contact,
%
the fraction of \cetis{} where the first contact is given at the
awakening,
%
the distribution of the number of contacts for each \ceti{},
%
the distribution of the number of contacts as a function of the age of
the node,
%
the number of contacts as a function of time in the galaxy,
%
the fractions of nodes that succeed in making contact,
%
and the distribution of distances between contacted nodes.
%
Another useful derived quantity is the duration of two-way
communication channels or the fraction of contacts that admit a
response.
% 
It is also possible to analyze the relations between the distance to
\ceti{} vs. the time of two-way communications, the distance to
\ceti{} vs. the age of contacted node, the age of a node and the
maximum number of contacted nodes before the D--type event, or the
lifespan of a node vs. the maximum number of contacts.
%
All these quantities can be analyzed as a function of the simulation
parameters.

@@

\ack[Acknowledgement]{
%
This work was partially supported by the Consejo Nacional de
Investigaciones Cient\'{\i}ficas y T\'ecnicas (CONICET, Argentina),
the Secretar\'{\i}a de Ciencia y Tecnolog\'{\i}a, Universidad Nacional
de C\'ordoba, Argentina, and the Universidad Cat\'olica de C\'ordoba,
Argentina.
%
This research has made use of NASA's Astrophysics Data System.
%
Plots and simulations were made with software developed by the authors
the python language. 
%
Plots were postprocessed with inkscape.
%
Simulations were run on the Clemente cluster at the Instituto de
Astronomía Teórica y Experimental (IATE).
%
}

@@

\bibliographystyle{mn2e}
\setlength{\bibsep}{0.0pt}
%\bibliography{biblio_seti, biblio_software}

\begin{thebibliography}{95}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

\bibitem[{Adamic(2000)}]{adamic_zipf_2000}
Adamic, L.A., 2000. Zipf, power-laws, and pareto-a ranking tutorial. Xerox Palo Alto Research Center, Palo Alto, CA, http://ginger. hpl. hp. com/shl/papers/ranking/ranking. html.

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{graphics}
\usepackage{algorithmic}
\usepackage{subfig}
\usepackage{url}
\usepackage{natbib}
\usepackage{boites}

@@

@@

@@

@@

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with
technologically advanced civilisations in the Milky Way. 
%
The equation quantifies the number of civilisations from whom we might
receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimated
values are revised continually. 
%
A comprehensive review and an analysis of each term of the equation
are presented in \citet{vakoch_drake_2015}.
%
Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the apparent contradiction
between the expected abundance of life in the Galaxy and the lack of
evidence for it \citep[e.g. ][]{hart_explanation_1975,
brin_great_1983, barlow_galactic_2012, forgan_galactic_2017,
anchordoqui_is_2017, Sotos_biotechnology_2019,
carroll_nellemback_fermi_2019}.
%
There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016,
horvat_calculating_2007, maccone_statistical_2015} or stochastic
approaches \citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}.
%
Regarding the Drake equation, analytical interpretations
\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations
\citep[][and references therein]{burchell_whither_2006} have also been
proposed. 
%
The absence of detections of extraterrestrial intelligent signals
could be explained by astrophysical phenomena that makes life
difficult to develop \citep{annis_astrophysical_1999}.
%
Besides the possible scarcity of life, alternative scenarios have also
been discussed \citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The large distances in the Galaxy and the likely limited lifetime of
civilisations may play an important role in determining how difficult
it would be to obtain evidence for other inhabited worlds.
%
The analysis of these scenarios is difficult due to the lack of data
about the hypothetical extraterrestrial intelligences.
%
Indeed, as \citet{tarter_search_2001} pointed out, according to our
current technical capabilities for the search of extraterrestrial
intelligence (SETI), we have not received any signal yet.
%
The absence of detections has also motivated alternative ideas for new
SETI strategies \citep{forgan_exoplanet_2019, balbi_impact_2018,
loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009,
enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011,
lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013,
maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The discussion about the problem of the unknown abundance of
civilisations in the Galaxy has been organised around the factors in
the Drake equation \citep{hinkel_interdisciplinary_2019}.
%
However, the uncertainties in these factors, specially the ones
representing biological processes, make it less suitable to a formal
study with the purpose of defining searching strategies or computing
the estimated number of extraterrestrial intelligences.
%
A number of studies propose alternative formalisms for the estimation
of the likelihood of detecting intelligent signals from space.
%
\citet{prantzos_joint_2013}, for example, proposes a unified framework
for a joint analysis of the Drake equation and the Fermi paradox,
concluding that for sufficiently long-lived civilisations,
colonisation is the most promising strategy to find other life forms.
%
\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host stars and the
time since the Galaxy formed, and examine trajectories for the
emergence of communicative civilisations.
%
Some modifications to the original idea of the Drake equation have
been proposed, in order to integrate a temporal structure, to
reformulate it as a stochastic process or to propose alternative
probabilistic expressions.
%
Temporal aspects of the distribution of communicating civilisations
and their contacts have been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balbi_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.

@@

On this topic a number of works consider numerical simulations
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017,
ramirez_new_2017}.
%
Although this approach does not rely on values that can be measured
from observations (like the fraction of stars with planets), it
depends on the definition of unknown or uncertain parameters required
to carry out the simulations.
%
In another numerical approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account the neighbour cells.
%
The authors implement this model to a network of cells which represent
life complexity on a two--dimensional region resembling the Galactic
Habitable Zone (GHZ), an annular ring set between a minimal radius of
6~kpc and a peak radius of 10~kpc.
%
These simulations represent the spread of intelligence as an
implementation of panspermia theories.
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
Monte Carlo simulations and analyse ensemble-averaged results.
%
Their work aims at analysing the evolution of life, and although it
does not account for the network of causal contacts among
technological civilisations, it offers a tool to think SETI from a
novel point of view.

This study is conceived as an introduction to a simple probabilistic
model and a numerical exploration of its parameter space.
%
With these tools, we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
This approach does not require asumptions about, e.g., the origin of
life, the development of intelligence or the formation of habitable
planets according to stellar type.
%
Instead, we assume monoparametric function families to model the
appearance of points in the disk of the Galaxy and over the time,
which we call \textit{nodes}.
%
These nodes represent the locations of ideal intelligent agents that
are able to receive and emit signals with perfect efficiency in all
directions.
%
Then we analyse the network of causally connected nodes limited to
have a maximum separation representing the maximum distance a signal
can travel with an intensity above a fixed threshold.
%
A node is causally connected to other nodes if it is within their
light cone.
%
We adopt as a definition that a light cone is the region in spacetime
within the surface generated by the light emanating from a given point
in space in a given period of time.
%
For example, two nodes separated by 100 light years which appear with
a difference in time of 80 years and last 50 years each, will be
causally connected for 30 years.
%
This connection is not, though, bidirectional.
%
In fact, the second node will see the first one for 30 years, but the
first node will never acknowledge the existence of the second.
%
An scheme of this example is shown if Fig.~\ref{F_example}.
%
This simple tought experiment exemplify the importance of the time
variable to analyse the structure of connections.
%
The model can be described with three parameters, namely, the mean
separation between the appearance of nodes, the mean lifetime of the
nodes, and the maximum separation to allow an effective causal
contact.
%
The model presented here comprises the stochastic networks of
constrained causally connected nodes, together with the three
parameters and additional hypotheses.
%
Hereafter, we refer to this model as the SC3Net model, standing for
''Stochastic Constrained Causally Connected Network``.

@@

Space--time diagrams, where time and space are represented on the
horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and space
axes are years and light years, respectively.
%
Panel (a) represents the region $R$ (shaded polygon) which is causally
connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance,
in which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Fig.~\ref{F_scheme}.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon.
%
This polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between
the A-type event (concave vertex) and the D-type event (convex
vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and
$D_2$).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$.
%
The receiver node E$_1$ can listen to signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen to signals from the
emitter node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

As part of this benchmark, we assume that the capacity to emit and
receive signals occur at the same time.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occur roughly at coincident epochs.
%
In the ideal setup, this would be equivalent to nodes that send
messages isotropically and scan the local skies on all directions with
a perfect efficiency.
%
Another essential assumption is that all \cetis{} use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
In such a system we compute probabilities of a random node making
contact with another node, i.e., they are not specific for the case of
contacts with the Earth.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is about few parsecs, given that the signal was sent to a
specific direction.
%
It is straightforward to propose and implement a distribution of
maximum distances, although this would increase the model complexity
at the cost of a larger uncertainty.

@@

\hline
\end{tabular}
\caption{
   %
   Definition of independent variables and adopted values for fixed
   parameters that are part of the simulation.  Variable parameters
   define the spatial and temporal structure of the process and the
   maximum reach of the messages.}
   %
\label{T_simu_hypotheses}
\end{table*}

@@

In order to make a reproducible project, we developed the tools that
allow to run the simulations and obtain the results shown in this
work.
%
The simulations were implemented on a Python-3 code, dubbed \hs{}
(Lares et al., in preparation), which is publicly accessible through
the GitHub platform\footnote{https://github.com/mlares/hearsay} under
the MIT-license.
%
The project is in the process of registration with the ‘’Astrophysics
Source Code Library’’ \citep[ASCL, ][]{2015JORS....3E..15A,
2020ASPC..522..731A}
%
From the viewpoint of a user, \hs{} is an object-oriented package that
exposes the main functionalities as classes and methods.
%
The code fulfills standar quality assurance metrics, that account for
testing, style, documentation and coverage.
%
In the configuration step, the user prepares the set of simulation
parameters through an initialisation file.
%
Since configuration files and simulation results are persistent, it is
straightforward to keep track of different experiments and the
experiments can be revisited easily.
%
An in-depth description of the methods can be found in the
documentation, which is automatically generated from \hs{} docstrings
and made public in the read-the-docs service
\footnote{https://hearsay.readthedocs.io/en/latest/}.
%
Since the simulation setup is configurable, the time required for a
simulation to complete depends on the simulation parameters.
%
It also depends on the hardware that is used to make the run, and on
whether the parallel option is set.
%
However, it is simple to make a local run of limited versions of the
experiments to have a sense of the time required by the code to
complete the simulations.
%
More information on this can be found on the documentation of the
software.

@@

We implemented the simulation of a regular grids of models varying
over the parameter space, which covers 5204 models.
%
For each model, we simulated several realizations with different
random seeds, adding up a total of 158546 simulation runs.
%
The number of random realzations varies from one (for the densest
populated models) up to 50 (for the sparsely populated models).
%
The parameters for the temporal aspects of the simulation (the mean
waiting time for the next awakening, $\tau_{\rm a}$, and the mean
lifetime, $\tau_{\rm s}$) cover the ranges 10$^2$-10$^6$~yr, with two
linear partitions of 21 values (10$^2$-40500~yr) and 51 values
($5\cdot10^5$-10$^6$~yr) for each parameter.
%
This partition was chosen after the requirement of the software to
take linear bins, aiming at a better sampling of the low $\tau_{\rm
a}$ and low $\tau_{\rm s}$ region of the parameter space.
%
For the D$_{\rm max}$ parameter, we take the values 100~pc, 500~pc,
1~kpc, 5~kpc and 10~kpc.
%
In the Table \ref{T_simu_hypotheses} we show the three variable
parameters, the ranges of their values and the number of bins that
have been explored in the numerical experiments.
%
We also show the set of fixed parameters that take part in the
simulation, their values and the hypotheses that they represent.

As a product of the simulations, several quantities can be obtained.
%
Some of them are directly derived from the discrete events, namely,
the ID of emitting and receiving \cetis{} and the position in the
galaxy.
%
We also save the times of each of the events that are relevant to keep
track of the number of \cetis{} for each simulation, i.e., the times
of the four types of events.
%
The times of C-type and B-type events are used to derive the number of
contacts.
%
We also obtain quantities that represent the properties of the
\cetis{}, for example the total time elapsed between the A-type and
the D-type of each node, which represent their corresponding
lifetimes.
%
The time span of a \ceti{} listening another or being listened by
another node can also be derived by keepig track of the times of the
events in a simulation.
%
This way we can also compute the distribution in the simulated galaxy
of \cetis{} that reach contact, the distributions of the waiting times
until the first contact or the distributions of waiting times until
the next contact.
%
The following properties of the population of \cetis{} can also be
derived:
%
the fraction of the lifetime a node is listening to at least another
node (i.e., within their light cone), the age of contacted \cetis{} at
first contact, the fraction of \cetis{} where the first contact is
given at the awakening, the distribution of the number of contacts for
each \ceti{}, the distribution of the number of contacts as a function
of the age of the node, the number of contacts as a function of time
in the galaxy, the fractions of nodes that succeed in making contact,
and the distribution of distances between contacted nodes.
%
Another useful derived quantity is the duration of two-way
communication channels or the fraction of contacts that admit a
response.
% 
It is also possible to analyze the relations between the distance to
\ceti{} vs. the time of two-way communications, the distance to
\ceti{} vs. the age of contacted node, the age of a node and the
maximum number of contacted nodes before the D--type event, or the
lifespan of a node vs. the maximum number of contacts.
%
All these quantities can be analyzed as a function of the simulation
parameters.

@@

\ack[Acknowledgement]{
%
This work was partially supported by the Consejo Nacional de
Investigaciones Cient\'{\i}ficas y T\'ecnicas (CONICET, Argentina),
the Secretar\'{\i}a de Ciencia y Tecnolog\'{\i}a, Universidad Nacional
de C\'ordoba, Argentina, and the Universidad Cat\'olica de C\'ordoba,
Argentina.
%
This research has made use of NASA's Astrophysics Data System.
%
Plots and simulations were made with software developed by the authors
the python language. 
%
Plots were postprocessed with inkscape.
%
Simulations were run on the Clemente cluster at the Instituto de
Astronom\'{\i}a Te\'orica y Experimental (IATE).
%
}

@@

\begin{thebibliography}{90}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Adamic(2000)]{adamic_zipf_2000}
{Adamic, L.A.} 2000,
\newblock {Zipf, Power-laws, and Pareto - a ranking tutorial}.
\newblock Information Dynamics Lab, HP Labs
\newblock URL \url{https://www.hpl.hp.com/research/idl/papers/ranking/ranking.html}.

\bibitem[{Allen} \& {Schmidt}(2015)]{2015JORS....3E..15A}
\newblock {{Allen, A., Schmidt, J.}} 2015,
\newblock {{Looking before Leaping: Creating a Software Registry}}.
\newblock \textit{Journal of Open Research Software}, \textbf{3},  E15, 2015.
